Report on the Second Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE2)

### 2015-07-05 WSSSPE2_report/WSSSPE2_review.tex
undoing changes

%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\subsection{Papers}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
The papers that were discussed in the Exploring Sustainability group are:
\begin{itemize}
\item Mario Rosado de Souza, Robert Haines, and Caroline Jay. Defining
sustainability through developers' eyes: Recommendations from an interview
study~\cite{wssspe2_rosada_de_souza}

\item Robert Downs, W. Christopher Lenhardt, Erin Robinson, Ethan Davis, and
Nicholas Weber. Community recommendations for sustainable scientific
software~\cite{wssspe2_downs}

\item Abani Patra, Matthew Jones, Steven Gallo, Kyle Marcus, and Tevfik Kosar.
Role of online platforms, communications and workflows in developing sustainable
software for science communities~\cite{wssspe2_patra}

\item Marlon Pierce, Suresh Marru, and Chris Mattmann. {WSSSPE2}: Patching it up,
pulling it forward~\cite{wssspe2_pierce}

@@

\item Colin C. Venters, Michael K. Griffiths, Violeta Holmes, Rupert R. Ward, and
David J. Cooke. The Nebuchadnezzar effect: dreaming of sustainable software
through sustainable software architectures~\cite{wssspe2_venters}
\end{itemize}

@@

Group discussion started with a position statement by each person surrounding
what they had learned from their software development experiences and how those
lessons might be translated into actionable outcomes. Participants came from a
range of backgrounds, and represented multiple software and training
initiatives, including Software Carpentry (SwC), Data Carpentry (DC), Open Science for
Synthesis (OSS), the Community Surface Dynamics Modeling Systems group, ROpenSci,
DataONE, the HDF Group, and others. Some of these software experiences were
focused on development of new products for use in the sciences (e.g., ROpenSci,
CSDMS), and these recounted the difficulties of engaging with disciplinary
scientists in writing software. Software was clearly utilized broadly across the
various science disciplines represented, and it was developed within those
disciplines as well. Some researchers created software for statistical analysis
and modeling, while others used it to control instrumentation, collate data
across networks, collect information from respondents, and many other uses.
There was broad consensus that, within the disciplines represented, formal
training in any type of software development or engineering was rare among the
practitioners; most are self-taught, and develop software to get another job
done. Any ancillary utility of the software outside of the specific science
target was generally unplanned and few researchers would want to invest more
time to make their own software more re-usable.

@@

Three main actions were identified by Subgroup A that would be of interest to
participants and benefit the research community. These focused on the
desire to amplify the current community efforts in training and software
engineering by connecting the current training initiatives (e.g., SwC, DC, 
rOpenSci, OSS) that are of different durations at appropriate stages.
Generally it was felt that a modicum of interaction between short term workshops
(SwC, DC), medium term trainings (e.g., OSS), and longer-term courses (e.g.,
BIDS Data Science) would benefit from coordinating curriculum, discussing and
aligning prerequisites, and coordinating timing of courses. A training
coordination effort would go a long way towards amplifying the value of the
individual efforts and make them all more effective.

@@

\begin{itemize}
\item Michael R. Crusoe and C. Titus Brown. Channeling community contributions to
scientific software: A hackathon experience~\cite{wssspe2_crusoe}

\item Marcus Hanwell, Patrick O'Leary, and Bob O'Bara. Sustainable software
ecosystems: Software engineers, domain scientists, and engineers collaborating
for science~\cite{wssspe2_hanwell}

\item W. Christopher Lenhardt, Stanley Ahalt, Matt Jones, J. Aukema, S. Hampton,
S. R. Hespanh, R. Idaszak, and M. Schildhauer. {ISEES-WSSI} Lessons for
sustainable science software from an early career training institute on open
science synthesis~\cite{wssspe2_lenhardt}

\item Jory Schossau and Greg Wilson. Which sustainable software practices do
scientists find most useful?~\cite{wssspe2_schossau}

@@

\item Jordan Adams, Sai Nudurupati, Nicole Gasparini, Daniel Hobley, Eric
Hutton, Gregory Tucker, and Erkan Istanbulluoglu. Landlab: Sustainable software
development in practice ~\cite{wssspe2_adams}

\item Alice Allen and Judy Schmidt. Looking before leaping: Creating a software
registry~\cite{wssspe2_allen}

\item Carl Boettiger, Ted Hart, Scott Chamberlain, and Karthik Ram. Building
software, building community: lessons from the {ROpenSci} project~\cite{wssspe2_boettiger}

\item Yolanda Gil, Eunyoung Moon, and James Howison. No science software is an
island: Collaborative software development needs in
geosciences~\cite{wssspe2_gil}

\item Ted Habermann, Andrew Collette, Steve Vincena, Werner Benger, Jay Jay
Billings, Matt Gerring, Konrad Hinsen, Pierre de Buyl, Mark K\"{o}nnecke, Filipe
Rnc Maia, and Suren Byna. The Hierarchical Data Format ({HDF}): A foundation for
sustainable data and software~\cite{wssspe2_habermann}

\item Eric Hutton, Mark Piper, Irina Overeem, Albert Kettner, and James Syvitski.
Building sustainable software - the {CSDMS} approach~\cite{wssspe2_hutton}

\item James S. Spencer, Nicholas S. Blunt, William A. Vigor, Fionn D. Malone, W.
M. C. Foulkes, James J. Shepherd, and Alex J. W. Thom. The {H}ighly {A}ccurate
{N-DE}terminant ({HANDE}) quantum {Monte Carlo} project: Open-source stochastic
diagonalisation for quantum chemistry~\cite{wssspe2_spencer}
\end{itemize}

@@

This group, with just three papers but a high level of interest and
participation from attendees, focused on the institutional, social, and cultural
mechanisms that encourage the creation and maintenance of shared software, in the
context of what now exists, what mechanisms are desired, and how
we might achieve them.

@@

Third that we could change academic culture, rather than worrying about the
systems. This was mostly focused on citations, because they matter for hiring,
promotion, and tenure decisions. The group discussed how we could weigh the
citations within papers better than we now do. How could we identify the five
citations that really matter for a paper, distinguishing them from the related
works and general background that are also cited. Perhaps we could break these
out in the reference list, working with publishers to implement this. Or maybe
we could also break out categories of citations, such as the most important
software used, the previous publication that we are building from, the data that
we actually used, etc. The Moore Foundation's award in data science was given as
an example, asking proposers: What are the five canonical citations that are
most important to your work?~\cite{moore-canonical} This would be a way of
giving credit and assigning importance to these works, differently from how we
just count citations today. A possible action that the group discussed was
conducting a longitudinal study of most useful software, data, etc.\ in a
discipline.

@@

This group started by discussing who should be incentivized, thinking of two
categories of people: those in science (who could be incentivized to do better,
more shareable, more sustainable work), and those in industry but interested in
science (who could be incentivized to contribute to science). It was pointed out
that we are not yet clear enough on exactly what we want to incentivize,
suggesting that we need to have a clearer picture of ``good computational work''
and what sort of contributions are truly generative for science.

The discussion of incentivizing those in science acknowledged that the
publications system was far from perfect for incentivizing good software work.
Nonetheless, the discussion focused on bringing software people into publications.
There were two main suggestions. The first is to focus on end users of software
and encourage them to cite software the ``right'' way. James Howison suggested
that his research showed that few projects were making a formal request for
citation (but that authors were not necessarily following those suggestions
anyway)~\cite{howison2015jasist}. He suggested making access to the software
conditional on a license that requires citation. Others found this ``too
confrontational'' and preferred to concentrate on making it easier to do the
right thing. The second was focusing on those leading software projects, and the
group was more enthusiastic about ``forcing'' PIs to include their ``software
people'' on publications, although there were few ideas on how exactly to do
this. Another technique mentioned was that when scientific software projects are
hosted in organizations like Apache, the scientific contributors can benefit
from building their reputations, perhaps yielding job offers that they can use
to negotiate better job and career packages.

@@

In the remaining discussion sessions, the group chose to split into three
subgroups to discuss these topics: {\em citation ecosystems}, {\em
taxonomy of contributorship/guidelines for software for tenure review}, and {\em
prizes}. The subgroups were asked to clearly identify
\begin{itemize}
\item the problem to be solved
\item steps towards a solution.
\end{itemize}

>>>>

%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\subsection{Papers}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
The papers that were discussed in the Exploring Sustainability group are:
\begin{itemize}
\item Mario Rosado de Souza, Robert Haines, and Caroline Jay. Defining
sustainability through Developers' Eyes: Recommendations from an Interview
Study~\cite{wssspe2_rosada_de_souza}

\item Robert Downs, W. Christopher Lenhardt, Erin Robinson, Ethan Davis, and
Nicholas Weber. Community Recommendations for Sustainable Scientific
Software~\cite{wssspe2_downs}

\item Abani Patra, Matthew Jones, Steven Gallo, Kyle Marcus, and Tevfik Kosar.
Role of Online Platforms, Communications and Workflows in Developing Sustainable
Software for Science Communities~\cite{wssspe2_patra}

\item Marlon Pierce, Suresh Marru, and Chris Mattmann. {WSSSPE2}: Patching It Up,
Pulling It Forward~\cite{wssspe2_pierce}

@@

\item Colin C. Venters, Michael K. Griffiths, Violeta Holmes, Rupert R. Ward, and
David J. Cooke. The Nebuchadnezzar Effect: Dreaming of Sustainable Software
through Sustainable Software Architectures~\cite{wssspe2_venters}
\end{itemize}

@@

Group discussion started with a position statement by each person surrounding
what they had learned from their software development experiences and how those
lessons might be translated into actionable outcomes. Participants came from a
range of backgrounds, and represented multiple software and training
initiatives, including Software Carpentry, Data Carpentry, Open Science for
Synthesis (OSS), the Community Surface Dynamics Modeling Systems group, ROpenSci,
DataONE, the HDF Group, and others. Some of these software experiences were
focused on development of new products for use in the sciences (e.g., ROpenSci,
CSDMS), and these recounted the difficulties of engaging with disciplinary
scientists in writing software. Software was clearly utilized broadly across the
various science disciplines represented, and it was developed within those
disciplines as well. Some researchers created software for statistical analysis
and modeling, while others used it to control instrumentation, collate data
across networks, collect information from respondents, and many other uses.
There was broad consensus that, within the disciplines represented, formal
training in any type of software development or engineering was rare among the
practitioners; most are self-taught, and develop software to get another job
done. Any ancillary utility of the software outside of the specific science
target was generally unplanned and few researchers would want to invest more
time to make their own software more re-usable.

@@

Three main actions were identified by Subgroup A that would be of interest to
participants and benefit the research community. These focused on the
desire to amplify the current community efforts in training and software
engineering by connecting the current training initiatives (e.g., Software and Data Carpentry,
rOpenSci, OSS) that are of different durations at appropriate stages.
Generally it was felt that a modicum of interaction between short term workshops
(SwC, DC), medium term trainings (e.g., OSS), and longer-term courses (e.g.,
BIDS Data Science) would benefit from coordinating curriculum, discussing and
aligning prerequisites, and coordinating timing of courses. A training
coordination effort would go a long way towards amplifying the value of the
individual efforts and make them all more effective.

@@

\begin{itemize}
\item Michael R. Crusoe and C. Titus Brown. Channeling community contributions to
scientific Software: A Hackathon Experience~\cite{wssspe2_crusoe}

\item Marcus Hanwell, Patrick O'Leary, and Bob O'Bara. Sustainable software
ecosystems: Software Engineers, Domain Scientists, and Engineers Collaborating
for Science~\cite{wssspe2_hanwell}

\item W. Christopher Lenhardt, Stanley Ahalt, Matt Jones, J. Aukema, S. Hampton,
S. R. Hespanh, R. Idaszak, and M. Schildhauer. {ISEES-WSSI} Lessons for
Sustainable Science Software from an Early Career Training Institute on Open
Science Synthesis~\cite{wssspe2_lenhardt}

\item Jory Schossau and Greg Wilson. Which Sustainable Software Practices do
Scientists Find Most Useful?~\cite{wssspe2_schossau}

@@

\item Jordan Adams, Sai Nudurupati, Nicole Gasparini, Daniel Hobley, Eric
Hutton, Gregory Tucker, and Erkan Istanbulluoglu. Landlab: Sustainable Software
Development in Practice ~\cite{wssspe2_adams}

\item Alice Allen and Judy Schmidt. Looking before Leaping: Creating a Software
Registry~\cite{wssspe2_allen}

\item Carl Boettiger, Ted Hart, Scott Chamberlain, and Karthik Ram. Building
Software, Building Community: Lessons from the {ROpenSci} Project~\cite{wssspe2_boettiger}

\item Yolanda Gil, Eunyoung Moon, and James Howison. No Science Software Is an
Island: Collaborative Software Development Needs in
Geosciences~\cite{wssspe2_gil}

\item Ted Habermann, Andrew Collette, Steve Vincena, Werner Benger, Jay Jay
Billings, Matt Gerring, Konrad Hinsen, Pierre de Buyl, Mark K\"{o}nnecke, Filipe
Rnc Maia, and Suren Byna. The Hierarchical Data Format ({HDF}): A Foundation for
Sustainable Data and Software~\cite{wssspe2_habermann}

\item Eric Hutton, Mark Piper, Irina Overeem, Albert Kettner, and James Syvitski.
Building Sustainable Software - The {CSDMS} Approach~\cite{wssspe2_hutton}

\item James S. Spencer, Nicholas S. Blunt, William A. Vigor, Fionn D. Malone, W.
M. C. Foulkes, James J. Shepherd, and Alex J. W. Thom. The {H}ighly {A}ccurate
{N-DE}terminant ({HANDE}) Quantum {Monte Carlo} Project: Open-source Stochastic
Diagonalisation for Quantum Chemistry~\cite{wssspe2_spencer}
\end{itemize}

@@

This group, with just three papers but a large amount of interest and
participation from attendees, focused on the institutional, social, and cultural
mechanisms that encourage the creation and maintenance of shared software, in the
context of what now exists, what mechanisms are desired, and how
we might achieve them.

@@

Third that we could change academic culture, rather than worrying about the
systems. This was mostly focused on citations, because they matter for hiring,
promotion, and tenure decisions. The group discussed how we could weigh the
citations within papers better than we now do. How could we identify the five
citations that really matter for a paper, distinguishing them from the related
works and general background that are also cited? Perhaps we could break these
out in the reference list, working with publishers to implement this. Or maybe
we could also break out categories of citations, such as the most important
software used, the previous publication that we are building from, the data that
we actually used, etc. The Moore Foundation's award in data science was given as
an example, asking proposers: What are the five canonical citations that are
most important to your work?~\cite{moore-canonical} This would be a way of
giving credit and assigning importance to these works, differently from how we
just count citations today. A possible action that the group discussed was
conducting a longitudinal study of most useful software, data, etc.\ in a
discipline.

@@

This group started by discussing who should be incentivized, thinking of two
categories of people: those in science (who could be incentivized to do better,
more shareable, more sustainable work), and those in industry but interested in
science (who could be incentivized to contribute to science.) It was pointed out
that we are not yet clear enough on exactly what we want to incentivize,
suggesting that we need to have a clearer picture of ``good computational work''
and what sort of contributions are truly generative for science.

The discussion of incentivizing those in science acknowledged that the
publications system was far from perfect for incentivizing good software work.
Nonetheless, the discussion focused on bringing software people into publications.
There were two main suggestions. The first is to focus on end users of software
and encourage them to cite software the ``right'' way. James Howison suggested
that his research showed that few projects were making a formal request for
citation (but that authors weren't necessarily following those suggestions
anyway)~\cite{howison2015jasist}. He suggested making access to the software
conditional on a license that requires citation. Others found this ``too
confrontational'' and preferred to concentrate on making it easier to do the
right thing. The second was focusing on those leading software projects, and the
group was more enthusiastic about ``forcing'' PIs to include their ``software
people'' on publications, although there were few ideas on how exactly to do
this. Another technique mentioned was that when scientific software projects are
hosted in organizations like Apache, the scientific contributors can benefit
from building their reputations, perhaps yielding job offers that they can use
to negotiate better job and career packages.

@@

In the remaining discussion sessions, the group chose to split into three
subgroups to discuss a version of these topics: {\em citation ecosystems}, {\em
taxonomy of contributorship/guidelines for software for tenure review}, and {\em
prizes}. The subgroups were asked to clearly identify
\begin{itemize}
\item the problem to be solved
\item steps towards a solution.
\end{itemize}
